```{r}
knitr::opts_chunk$set(eval = FALSE)
library(dplyr)
library(ggplot2)
```

## Introduction

**Goal**: by the end of this lab, you will be able to use `ggplot2` to build several different data-driven graphics.

### Setting up
**Remember**: before we can use a library like `ggplot2`, we have to load it:

```{r}
library(ggplot2)
```


### Why `ggplot2`?

Advantages of `ggplot2`

-   consistent underlying **grammar of graphics** (Wilkinson, 2005)
-   plot specification at a high level of abstraction
-   very flexible
-   `theme` system for polishing plot appearance (more on this later)
-   mature and complete graphics system
-   many users, active mailing list

### What Is The Grammar Of Graphics?

**The big idea**: independently specify plot building blocks and combine them to create just about any kind of graphical display you want. Building blocks of a graph include:

-   data
-   aesthetic mappings
-   geometric objects
-   statistical transformations
-   scales
-   coordinate systems
-   position adjustments
-   faceting

Using `ggplot2`, we can specify different parts of the plot, and combine them together using the `+` operator.

### Example Data: `Housing prices`

Let's start by taking a look at some data on housing prices:

``` {r}
housing <- read.csv("https://amosca01.github.io/SSEP-22/data/landdata-states.csv", 
                    header = T)
glimpse(housing)
```

(Data from <https://www.lincolninst.edu/subcenters/land-values/land-prices-by-state.asp>)

## Geometric Objects and Aesthetics

### Geometric Objects (`geom`)

Geometric objects or `geoms` are the actual marks we put on a plot. Examples include:

-   points (`geom_point`, for scatter plots, dot plots, etc)
-   lines (`geom_line`, for time series, trend lines, etc)
-   boxplot (`geom_boxplot`, for, well, boxplots!)
-   ... and many more!

A plot must have at least one `geom`; there is no upper limit. You can add a `geom` to a plot using the `+` operator.

You can get a list of available geometric objects by looking at the package documentation, available through the help interface:

``` {r}
?ggplot2
```

or simply type `geom_<tab>` in any good R IDE (such as Rstudio) to see a list of functions starting with `geom_`.

### Aesthetic Mapping (`aes`)

In `ggplot2`-land, *aesthetic* means "something you can see". Examples include:

-   position (i.e., on the x and y axes)
-   color ("outside" color)
-   fill ("inside" color)
-   shape (of points)
-   line type
-   size

Each type of `geom` accepts only a subset of all aesthetics -- refer to the `geom` help pages to see what mappings each `geom` accepts. Aesthetic mappings are set with the `aes()` function, short for aesthetic. 


### Points

Now that we know about geometric objects and aesthetic mapping, we're ready to make our first `ggplot`: a scatterplot. We'll use `geom_point` to do this, which requires `aes` mappings for `x` and `y`; all others are optional. 


``` {r}
hp2013Q1 <- housing %>% filter(Date == 2013.25) 

ggplot(hp2013Q1,
       aes(x = Land.Value,
           y = Structure.Cost)) +
  geom_point()
```
### Plot objects

The output of the `ggplot()` function is an object. If we want to modify the plot that we created above, it’s helpful to store the plot as an object.

```{r}
p1 <- ggplot(hp2013Q1, 
             aes(x = Land.Value,      
                 y = Structure.Cost)) +
  geom_point()
```

To actually show the plot, we just print it. Note: the aesthetic mappings defined in p1, will be used by any subsequent `geom`s that we add unless we override them.

```{r}
p1
```

### Lines

A plot constructed with `ggplot` can have more than one `geom`. In that case, the mappings established in the `ggplot()` call are plot defaults that can be added to or overridden. For example, we could add an aesthetic mapping for color to `Home.Value`:

``` {r}
p1 + 
  geom_point(aes(color = Home.Value)) 
```

And we could add a regression line to our plot:
``` {r}
p1 + 
    geom_smooth(method = "lm")
```

### Smoothers

Not all geometric objects are simple shapes -- as we saw above, `geom_smooth` includes both a line and a ribbon.

``` {r}
p1 +
  geom_point(aes(color = Home.Value)) +
  geom_smooth() # If we don't specify the method, what happens? 
                # Check the docs for more details (?geom_smooth)
```

### Removing standard errors

If you don't need the error bands, just set `se = FALSE` inside the call to `geom_smooth` to remove them:
``` {r}
p1 +
  geom_point(aes(color = Home.Value)) +
  geom_smooth(se = FALSE)
```

### Text

Each `geom` accepts a particualar set of mappings--for example `geom_text()` accepts a `labels` mapping.

``` {r}
p1 + 
  geom_point(aes(color = Home.Value)) +
  geom_text(aes(label = State), size = 3)
```

#. Try playing around with the `position` or `nudge_x/nudge_y` arguments to `geom_text` to see if you can offset the label from the associated point a bit.

### Aesthetic Mapping VS Assignment

Note that variables are mapped to aesthetics with the `aes()` function, while fixed visuals are set outside the `aes()` call. This sometimes leads to confusion, as in this example:

``` {r}
p1 +
  geom_point(aes(size = 2),     # not quite right! 2 is not a variable, but rather a fixed value
                 color = "red") # this is fine -- turns all points red
```
Notice how we get a pretty useless legend for size? That's because `R` is interpreting the constant `2` as a variable (just a boring one containing a single value) and so produces a legend by default.

#. Modify the code above to get rid of the legend by assigning the `size` attribute outside the call to `aes()`. While you're at it, try mapping `region` to `color` to see if there are any interesting patterns!

### Mapping Variables to Other Aesthetics

Other aesthetics are mapped in the same way as `x` and `y` in the previous example.

``` {r}
p1 +
  geom_point(aes(color = region, 
                 shape = region),
             size = 5)
```


## Scales: Controlling Aesthetics

Aesthetic mapping (i.e., with `aes()`) only says that a variable should be mapped to an aesthetic. It doesn't say *how* that should happen. For example, when mapping a variable to *shape* with `aes(shape = myVar)` you don't say *which* shapes should be used. Similarly, `aes(color = z)` doesn't say *which* colors should be used. Describing what colors/shapes/sizes etc. to use is done by modifying the corresponding *scale*. In `ggplot2`, `scales` include:

-   `position`
-   `color` and `fill`
-   `size`
-   `shape`
-   `linetype`

#.  Scales are modified with a series of functions using a `scale_<aesthetic>_<type>` naming scheme. Try typing `scale_<tab>` to see a list of scale modification functions, then play around to make the plot above look a little nicer.

### Common Scale Arguments

The following arguments are common to most `scales` in `ggplot2`:

-   `name`:  the first argument specifies the axis or legend title
-   `limits`: the minimum and maximum of the `scale`
-   `breaks`:  the points along the scale where labels should appear
-   `labels`: the text that appear at each break

#. Specific scale functions may have additional arguments; for example, the `scale_color_continuous` function has arguments `low` and `high` for setting the colors at the low and high end of the scale. Try playing around with the color scale to keep the outliers from dominating in this plot:

```{r}
# Tweak the color scale!
p1 + 
  geom_point(aes(color = Home.Value))
```

### Scale Modification Examples

Start by constructing a dotplot showing the distribution of home values by `Date` and `State`.

``` {r}
p3 <- ggplot(housing,
             aes(x = State,
                 y = Home.Price.Index))

p4 <- p3 + geom_point(aes(color = Date),
                      alpha = 0.5,
                      size = 1.5,
                      position = position_jitter(width = 0.25, height = 0))

p4
```
First, we will change the label on the vertical axis.

```{r}
p4 <- p4 + 
  scale_x_discrete(name = "State Abbreviation")

p4
```

Now let's modify the `breaks` and `labels` for the `x` axis and color scales:

``` {r}
p4 <- p4 + 
  scale_color_continuous(name = "",
                          breaks = c(1975.25, 1994.25, 2013.25),
                          labels = c(1971, 1994, 2013))
p4
```

Next change the low and high values to `blue` and `red`:

``` {r}
p4 <- p4 +
  scale_color_continuous(name = "",
                          breaks = c(1975.25, 1994.25, 2013.25),
                          labels = c(1971, 1994, 2013),
                          low = "blue", high = "red")
p4
```

### Using different color scales

`ggplot2` has a wide variety of color `scales`; here is an example using `scale_color_gradient2` to interpolate between three different colors:

``` {r}
p4 +
  scale_color_gradient2(name = "",
                        breaks = c(1975.25, 1994.25, 2013.25),
                        labels = c(1971, 1994, 2013),
                        low = "blue",
                        high = "red",
                        mid = "gray60",
                        midpoint = 1994.25)
```


### Available Scales

Here's a (partial) combination matrix of available scales:

| **Scale**         | **Types**  | **Examples**                     |
|-------------------|------------|----------------------------------|
| `scale_color_`    | `identity`   | `scale_fill_continuous`   |
| `scale_fill_`     | `manual`     | `scale_color_discrete`    |
| `scale_size_`     | `continuous` | `scale_size_manual`       |
|                            | `discrete`   | `scale_size_discrete`     |
| `scale_shape_`    | `discrete`   | `scale_shape_discrete`    |
| `scale_linetype_` | `identity`   | `scale_shape_manual`      |
|                            | `manual`     | `scale_linetype_discrete` |
| `scale_x_`        | `continuous` | `scale_x_continuous`      |
| `scale_y_`        | `discrete`   | `scale_y_discrete`        |
|                            | `reverse`    | `scale_x_log`             |
|                            | `log`        | `scale_y_reverse`         |
|                            | `date`       | `scale_x_date`            |
|                            | `datetime`   | `scale_y_datetime`        |

**Note**: in RStudio, you can type `scale_` followed by TAB to get the whole list of available scales.

## Faceting

-   Faceting is `ggplot2` parlance for **small multiples**
-   The idea is to create separate graphs for subsets of data
-   `ggplot2` offers two functions for creating small multiples:
    -  `facet_wrap()`: define subsets as the levels of a single grouping variable
    -  `facet_grid()`: define subsets as the crossing of two grouping variables
-   Facilitates comparison among plots, not just of `geoms` within a plot

### Example: what is the trend in housing prices in each state?

Let's start by using a technique we already know -- map `State` to `color`:

``` {r}
p5 <- ggplot(housing, aes(x = Date, 
                          y = Home.Value))

p5 + geom_line(aes(color = State))  
```

There are two problems here--there are too many `states` to distinguish each one by `color`, and the `lines` obscure one another.

### Faceting to the rescue!

We can fix the previous plot by `faceting` by `State` rather than mapping `State` to `color`:

``` {r}
p5 + geom_line() +
    facet_wrap(~State, ncol = 10)
```

There is also a `facet_grid()` function for faceting in two dimensions.


## Remember Tidy Data? 

### Map 2 (or more) Different Columns to one Aesthetic

A common question from `ggplot` beginners goes something like this: *I have two variables in my dataframe, and I'd like to plot them as separate points, with different colors depending on which variable it is. How do I do that?*

### First try: draw two separate lines

``` {r}
housing_byyear <- aggregate(cbind(Home.Value, Land.Value) ~ Date, data = housing, mean)

ggplot(housing_byyear, aes(x=Date)) +
  geom_line(aes(y = Home.Value), color="red") +
  geom_line(aes(y = Land.Value), color="blue")
```

This works but it is not so great, because if I later add another dimension, I'll have to remember that I hard-coded these lines separately. Plus because I didn't use an `aes()` mapping, I don't get a legend. Wah wah.

### Second try: A tidier way

First we'll `pivot_longer` the values we want to plot:
``` {r}
library(tidyr)
home_land_byyear <- pivot_longer(housing_byyear,
                            values_to = "value",
                            names_to = "type",
                            cols = c("Home.Value", "Land.Value"))
```

Then we'll plot them:
``` {r}
ggplot(home_land_byyear,
        aes(x = Date,
            y = value,
            color = type)) +
  geom_line()
```

Much cleaner!

## Your turn!
Now let's try building a few visualizations of your own. Try answering these questions using the `housing` dataset we've been using so far.

#. Create a scatterplot of the value of each home in the first quarter of 2013 as a function of the value of the land.
```{r}
# sample solution
    ggplot(hp2013Q1, aes(y = Home.Value, x = Land.Value)) + 
      geom_point()
```
#. Store the scatterplot that you created in the previous exercise as an object called `home_value_plot`.
```{r}
# sample solution
      home_value_plot <- ggplot(hp2013Q1,
      aes(y = Home.Value, x = Land.Value)) + 
    geom_point()
```
#. In your `home_value_plot`, map color to the cost of the structure and show your scatterplot.
```{r}
# sample solution
      home_value_plot + 
      aes(color = Structure.Cost) + 
      geom_point()
```
#. Let's return to p4. Since a home price index of 1 is an important benchmark, it is worth
highlighting as contextual reference in our plot. Use `geom_vline()` to add a dotted, black, vertical line to p4.
```{r}
# sample solution
p4 + 
      scale_color_gradient2(breaks = c(1975.25, 1994.25, 2013.25),
                        labels = c(1971, 1994, 2013),
                        low = "blue",
                        high = "red",
                        mid = "gray60",
                        midpoint = 1994.25) + 
      geom_hline(yintercept = 1, linetype = 3, color = "black")
```
#. Recall that layers in ggplot2 are added sequentially. How would you put the dotted horizontal line you created in the previous exercise **behind** the data values?
```{r}
# sample solution
p4 + 
  geom_hline(yintercept = 1, linetype = 3, color = "black") + 
  scale_color_gradient2(breaks = c(1975.25, 1994.25, 2013.25),
                        labels = c(1971, 1994, 2013),
                        low = "blue",
                        high = "red",
                        mid = "gray60",
                        midpoint = 1994.25) 
```

#. Use a `facet_wrap` to create a data graphic of your choice that illustrates something interesting about home prices.
```{r}
# sample solution
    ggplot(housing, aes(x = Date, y = Home.Price.Index, color = State)) + 
      geom_hline(aes(yintercept = 1), linetype = 3, color = "black") + 
      geom_point(alpha = 0.1) +
      geom_line() + 
      facet_wrap(~region)
```

<div id="license">
This lab is based on the "[Introduction to R Graphics with `ggplot2`](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html)" workshop, which
is a product of the [Data Science Services](http://dss.iq.harvard.edu/) team [Harvard University](http://dss.iq.harvard.edu/). The original source is released under a 
[Creative Commons Attribution-ShareAlike 4.0 Unported](http://creativecommons.org/licenses/by-sa/4.0). 
This lab was adapted for SDS192: an Introduction to Data Science in Spring 2017 by R. Jordan Crouser at Smith College, and for the MassMutual Data Visualization workshop in Summer 2017 by Amelia McNamara, with minor revisions in Summer 2020 by R. Jordan Crouser.
</div>