```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

In this lab, we will look at the entire analysis pipeline, from question generation to interpreting analysis results. 

```{r, message=FALSE}
library(tidyverse)
library(babynames)
```

**Goal**: by the end of this lab, you will be able to scope and perform your own analysis. 

## Setup

In the `babynames` package, there is a table called `births`. It has just two columns:

```{r}
head(births)
```


## Your turn!
Now let's try joining some tabvles from a dataset of your choosing. Don't have one? Try answering these questions using the `nycflights13` package.

#. Compute the average arrival delay time for each carrier, and include the full name of the carrier in your result set.

#. What was the full name of the airport that was the most common destination from NYC in 2013? 

#. What is the most commonly used plane? Find its manufacturer and model.

#. Were there any flights that went to "mystery" airports (i.e., airports that don't appear in the `airports` table)? 

#. Were there any "mystery" planes (i.e., planes that don't appear in the `planes` table)? 