```{r global_options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

In this lab, we will learn how to convert data tables into a tidy form. We will learn about functions in the `tidyr` package, which is part of the `tidyverse` meta-package. 

```{r, message=FALSE}
library(tidyverse)
library(babynames)
```

**Goal**: by the end of this lab, you will understand what tidy data is and how to convert a wide table into a narrow table, and vice versa. 

## What does `tidy` mean, and why do we care?

**Tidy** data are usually long and narrow. That is, it may have many rows, but relatively few columns. However, data are often stored in spreadsheets in a **wide** format (which may have more columns that rows) because the wide format is easier to **see** in a spreadsheet.

However, it turns out that there are lots of good reasons to keep data in a `tidy` format:

#. Tools like `dplyr` and `ggplot2` expect tidy data. 
#. It is basically the [Third Normal Form](https://en.wikipedia.org/wiki/Third_normal_form), which is the standard in relational databases (more on that later). 
#. Now that we know how to wrangle and make simple visualizations, we don't need to see it all anyway. 

## `tidyr`

The `pivot_longer()` and `pivot_wider()` functions from the `tidyr` package are **huge** time-savers. (I don't know of any equivalent operation in Excel that can do what these functions do. Just imagine what you'd have to do in Excel? ::shudder::) 

To recap:

- `pivot_longer()` takes a data table in a **wide** format and converts it into a tidy format. 
- `pivot_wider()` does the opposite (and accordingly, is used less frequently). 

## An example: making wide tables narrow with `pivot_wider()`
We (unwittingly) came up with an example yesterday. Recall that the `total_births` data frame looks like this:

```{r}
ssa_births <- babynames %>%
  group_by(year) %>%
  summarize(num_rows = n(), births = sum(n))

total_births <- births %>%
  left_join(ssa_births, by = "year")

head(total_births)
```

Let's forget about `num_rows` and rename the variables.

```{r}
total_births <- total_births %>%
  select(-num_rows) %>%
  rename(census = births.x, ssa = births.y)
```

Each row corresponds to a single `year`, and for each year, we have two variables: `census` (from the Census) and `ssa` (from the SSA). This data frame is **not** tidy. (Why?) 

This data frame has 109 rows and 3 columns:

```{r}
dim(total_births)
```

...but what we want is a data frame that has 218 rows and three columns: `year`, `source`, and `births`. This data frame would be tidy, because like is stored with like. The `pivot_longer()` function will help us get there.  

```{r}
births_tidy <- total_births %>%
  pivot_longer(cols = c(ssa, census), 
               names_to = "source", 
               values_to = "births") %>%
  arrange(year)

head(births_tidy)

dim(births_tidy)
```

See what happened? View `births_tidy` in several different ways if necessary, and make sure that you understand where each piece of data went. 

Foreshadowing what we'll learn about data visualization in a couple of weeks, we can now map `year` to the x axis, `births`= to the y axis, and use color to encode the `source`:

```{r}
ggplot(data = births_tidy, aes(x = year, y = births, color = source)) +
  geom_line() 
```

Note that even though we have two sources for birth numbers, the numbers themselves represent the same kind of thing. 

### Now, with a little less hand-holding...

Try this process out for yourself with a [Gapminder dataset](http://www.gapminder.org/data/). We'll walk you through it step-by-step using the "Prevalence of HIV, total (\% of population ages 15-49)" table, but feel free to choose any table that's interesting to you.

#### Step 1:
> Pick a dataset and download it as a CSV file.

#### Step 2:
> Read the CSV into R using `read_csv()`. Pay careful attention to the path names! Your code will look something like this:

```{r}
hiv <- read_csv("~/Downloads/sh_dyn_aids_zs.csv")
```

#### Step 3:

> Get to know your data. Look at the first few rows. Use `dim()` to find the dimensions of your data frame. Make a note of these. 

```{r}
head(hiv)
dim(hiv)
```

#### Step 4:

> Take out a piece of paper and draw what you want your "tidy" data frame to look like. It should have three columns: `country`, `year`, and `value`. How many rows will it have? 

#### Step 5:

> Use `pivot_longer()` and/or `pivot_wider()` to convert your data into a tidy format. Check that it has the right dimensions. 

```{r}
hiv <- hiv %>%
  pivot_longer(-country, 
               names_to = "year", 
               values_to = "est_prevalence") %>%
  mutate(year = sub("X", "", year))
```

#### Step 6:

> Say something interesting! Here's one example:

```{r}
hiv %>%
  ggplot(aes(x = as.numeric(year), 
             y = est_prevalence, 
             group = country,
             color = est_prevalence)) +
  geom_line() +
  xlim(c(1990, 2011))
```

